---
title: "Bayesian Latent Severity LCA"
output: rmarkdown::github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.path = "README_files/"
)
```


A **Bayesian latent class model with continuous severity** for diagnostic test evaluation.

This project extends classical Hui–Walter and random-effects LCA frameworks by introducing a continuous *latent severity* parameter \(S_i\).  
The approach enables efficient estimation of diagnostic accuracies and prevalence using **fewer than four assays**, addressing common identifiability limitations in small-panel diagnostic evaluations.


---

## Overview

Traditional latent class models assume conditional independence between binary tests given a latent disease status.  
The **Latent Severity LCA (S-LCA)** relaxes this by allowing within-class variability in disease expression:

\[
T_{ij}^* = \beta_j S_i + \gamma_j + \varepsilon_{ij}, 
\qquad T_{ij} = \mathbb{I}(T_{ij}^* > 0)
\]

Here, \(S_i\) captures the strength or intensity of infection for diseased subjects (\(D_i = 1\)), producing more realistic inference on test sensitivity, specificity, and prevalence.

---


```{r, echo = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
sample_NM <- function(n, mu0, tau, grid_size = 1e4) {
  x_vals <- seq(0, mu0 + 6 * tau, length.out = grid_size)
  dens_vals <- 2 * (x_vals^2) / (sqrt(2 * pi) * tau^3) * 
    exp(-((x_vals - mu0)^2) / (2 * tau^2))
  dx <- x_vals[2] - x_vals[1]
  dens_vals <- dens_vals / sum(dens_vals * dx)
  
  cdf_vals <- cumsum(dens_vals * dx)
  
  # Sample from uniform and interpolate inverse CDF
  u <- runif(n)
  samples <- approx(cdf_vals, x_vals, xout = u, rule = 2)$y
  
  return(samples)
}
N = 5000
vart = 1/sqrt(3 - 8/pi)
Si_nm  <- sample_NM(N, mu0 = 0, tau = vart)          
Si_gam <- rgamma(N, shape = 4.5, rate = sqrt(4.5))                

df <- tibble(
  NMplus = Si_nm,
  Gamma  = Si_gam
) |> 
  pivot_longer(everything(), names_to = "Distribution", values_to = "Si") |>
  mutate(
    Distribution = factor(
      Distribution,
      levels = c("NMplus", "Gamma"),
      labels = c("NM+ (μ=0, τ=1)", "Gamma (shape=4.5, rate=2.5)")
    )
  )

ggplot(df, aes(x = Si, fill = Distribution, color = Distribution)) +
  geom_density(alpha = 0.3, adjust = 1.2) +
  scale_x_continuous(limits = c(0, 8)) +
  labs(
    title = "Candidate priors for latent severity S",
    x = expression(S[i]), y = "Density"
  ) +
  theme_minimal(base_size = 13) +
  theme(legend.title = element_blank())
```

